Extensible Gradual Typing Inside a Python

### 2014-10-14 atlang-modularity15/atlang-modularity15.tex
updated submission to modularity due to extension

\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
%\usepackage{amsthm}
%\newtheorem{corollary}{Corollary}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
 \usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{pdflscape}
 \usepackage{colortbl}%
 \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
 \usepackage{wasysym}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, cls, @classmethod, with},
  deletekeywords={min,max},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  numbersep=0.5em,
  xleftmargin=1.5em,
  stringstyle=\color{mauve},
  escapeinside={^}{^},
  aboveskip=1pt,
  belowskip=1pt
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`}
}
\lstdefinestyle{OpenCL}{
  language=C++,
  morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small]{#1}}

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests that a language is needed that is {backwards compatible} with the libraries and tools of a widely adopted language, but that has an  {internally extensible type system}, organized around a mechanism that guarantees that any combination of extensions can be imported safely.

Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library inside Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax but  determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even when additional extensions are imported, so we avoid ambiguity by construction. We describe several examples of interesting type system fragments as libraries, including functional constructs, a type system for secure string sanitation and a foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several foreign languages often requires programming against a low-level internal implementation, rather than its semantics, via a foreign-function interface (FFI). This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries. As more languages are used together,  more FFIs must be developed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues can prevent even dialects introducing only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct control over a language's syntax and semantics and supporting the composition of these language fragments within or alongside libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

@@

Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing, widely-used language. The most immediately difficult problems related to composing type system fragments arise when attempting to enable the decentralized introduction of new concrete syntax or corresponding term constructors to support the operations of this new type system. Syntactic ambiguities are difficult to modularly preclude (though some techniques exist). Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Fewer tools operate by exhaustive case analysis over type constructors. In a conventionally formulated programming language, however, it is difficult to introduce a new type system fragment without adding new concrete syntax or term constructors to accompany the new type constructors. In this paper, we aim to show that this need not be the case. We introduce a method that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete syntax, the term constructors and the target language of compilation fixed. The key idea is that we determine the semantics of each term not based directly on the identity of its term constructor (the usual ``syntax-directed'' approach), but rather by delegating to a statically evaluated function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label). A fragment defining statically typed pairs might want this to refer to projection of the right component of the pair, while a fragment defining functional record types might want it project out the field labeled \verb|r| and an object system might introduce a more complex semantics involving dynamic dispatch. Were we to naively ``open up'' the syntax and semantics of the language, these fragments would conflict.

Instead, we assume that the syntax is fixed and includes this commonly used form, mapping it to abstract term  $\FF{attr}[\text{``r''}](e)$ (i.e. the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$). Our semantics delegate responsibility for this term to a static function associated with the type constructor of the type recursively synthesized by the ``target'' of the operation, $e$. Tuples, records, objects and variants thereof can all be supported over a common syntax. As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against (consistent with how extensible parsing is done in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction. %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

@@

%
Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which strong correctness guarantees have been statically maintained using several type constructors defined in libraries. As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python. @\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy). This choice pays off immediately on the first line: @\texttt{lang}'s import mechanism is Python's import mechanism, so Python's build tools (e.g. \lstinlinep{pip}) and package repostories (e.g. \lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries, including those defining type constructors. Here, \lstinlinep{atlib} is the ``standard library'' but does not benefit from special support in \lstinlinep{atlang} itself.% Nothing in \lstinlinep{atlib} benefits from special support from \lstinlinep{atlang}, the core language package. We will see \lstinlinep{atlang} being used directly below. %@\texttt{lang} can be installed with a single command: \emph{omitted for review}. %@\texttt{lang} is agnostic about the particular design choices that we made in our examples.

@@

\paragraph{Types}
Types are constructed programmatically in the compilation script by applying a type constructor to an index:  \lstinlinep{tycon[idx]}. This is in contrast to many statically-typed languages where types (e.g. datatypes, classes) are not constructed programmatically, but can only be declared ``literally''. %Put another way, 
%@\texttt{lang} supports \emph{static computation at the level of types} and Python is its type-level language. The static language subsumes th%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \lstinlinep{typedef} in C or \lstinlinep{type} in Haskell). 
In this example, we see several useful types being constructed:
\begin{enumerate}
\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, assigning it to the static identifier \lstinlinep{Details}. The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, classifying dynamic Python values.  %Here, \lstinlinep{record} and \lstinlinep{decimal} are {indexed type constructors}. 
Syntactically, the index for \lstinlinep{record} is provided using borrowed Python  syntax for array slices (e.g. \lstinlinep{a[start:end]}) to approximate conventional notation for type annotations. The field names are simply static strings (and could be computed rather than written literally, e.g. \lstinline{'a' + 'mount'}, to support features like \emph{type providers} \cite{Syme:2013:TIF:2429376.2429378}). 
\item On line \ref{decl-Account-start}, we construct another record type. The field \lstinlinep{name} has type \lstinlinep{string} and the field \lstinlinep{account_num} has a type classifying strings guaranteed statically to be in the regular language specified  by the  regular expression pattern provided as the index (written here as a \emph{raw string literal} \cite{python}). This  fragment, based on recent work, statically tracks the language an immutable string is in, providing operations like concatenation, demonstrating the expressive power of our approach \cite{sanitation-psp14}.%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line \ref{decl-Transfer}, we construct a simple \emph{prototypic object type} \cite{Lie86}, \lstinlinep{Transfer}, classifying values consisting of a \emph{fore} of type \lstinlinep{Details} and a \emph{prototype} of type \lstinlinep{Account}. %\footnote{We encourage the convention of naming types that are semantic in camelcase and types and type constructors that are structural in nature, like \texttt{dyn} or \texttt{record}, in lowercase.} 
 If a field cannot be found in the fore, the type system will delegate (statically) to the type of the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers to the same account.% differing only in the transaction  details.
\end{enumerate}

\paragraph{Active Type Constructors} Type constructors are implemented in @\texttt{lang} libraries as Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums). We show portions of \lstinlinep{atlib.fn} in Listings \ref{fn-top} and \ref{fn-stmts}, \lstinlinep{atlib.record} in Listing \ref{record} and \lstinlinep{atlib.string_in} in Listing \ref{string_in}, which we will discuss as we go on. The types in our example are instances of these classes.

\paragraph{Incomplete Types} Incomplete types arise from the application of a type constructor to an index containing an ellipsis (a valid array slice form in Python). The static terms \lstinlinep{fn[Transfer, ...]} and \lstinlinep{fn[...]} seen on lines \ref{log-transfer-start} and \ref{main-start} are incomplete types. We will discuss how the elided portions of an index (e.g. the return types, which need not be provided) are inferred below.

Incomplete types are not instances of \verb|atlang.Type| but rather instances of \lstinlinep{atlang.IncompleteType}. An incomplete type is defined by a type constructor (here \lstinlinep{fn}) and an incomplete index (here, \lstinlinep{(Transfer, Ellipsis)}).  The language controls instantiation by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).

@@

 Ascribed function definitions do not share Python's semantics and indeed the underlying Python function is discarded immediately after its abstract syntax and static environment (its closure and the global dictionary of the Python module it was defined in) have been extracted by the compiler using Python's built-in reflection capabilities and \lstinlinep{inspect} and \lstinlinep{ast} packages \cite{python}. The \lstinlinep{ast} package defines Python's term constructors.

>>>>

\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
%\usepackage{amsthm}
%\newtheorem{corollary}{Corollary}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
 \usepackage{longtable}
 \usepackage{booktabs}
 \usepackage{pdflscape}
 \usepackage{colortbl}%
 \newcommand{\myrowcolour}{\rowcolor[gray]{0.925}}
 \usepackage{wasysym}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, cls, @classmethod, with,as},
  deletendkeywords={tuple,id,buffer,map},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  numbersep=0.5em,
  xleftmargin=1.5em,
  stringstyle=\color{mauve},
  escapeinside={^}{^},
  aboveskip=1pt,
  belowskip=1pt
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`}
}
\lstdefinestyle{OpenCL}{
  language=C++,
  morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

\newcommand{\lstinlinep}[1]{\lstinline[language=Python,basicstyle=\ttfamily\small,deletendkeywords={tuple,buffer,map}]{#1}}

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce @\texttt{lang}, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, @\texttt{lang} is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, @\texttt{lang}'s type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in @\texttt{lang} can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers prefer using abstractions available as libraries for the language they already use over those requiring a switch to a new language. But library providers cannot typically extend the static type systems of these languages directly, so powerful statically typed abstractions  remain obscure. This suggests the need for a mechanism that allows abstraction providers to extend the type system of a widely adopted language from within, in a way that guarantees that any combination of extensions can be used together without worrying about conflicts.

Our aim in this paper is to 1) introduce such a mechanism; and 2) show how it can be implemented as a library for an existing widely used language, here Python. We describe two artifacts: @\texttt{lang}, an  extensible statically typed language implemented as a library within Python, and @$\lambda$, a minimal core calculus. Both @\texttt{lang} and @$\lambda$ have a fixed syntax. They determine the semantics of each term by delegating control, according to a simple but powerful bidirectionally typed elaboration semantics, to methods associated with a relevant \emph{active type constructor}, defined in a library. This process is stable even when additional extensions are imported, so we avoid syntactic and semantic ambiguity by construction. We validate our approach with several examples of interesting type system fragments that are built into other languages but can be implemented as libraries using our approach, including functional constructs, a type system for secure string sanitation and a typed foreign function interface to OpenCL.% By flexibly  repurposing a fixed syntax in this type-directed manner, we avoid ambiguities by construction. %In @\texttt{lang}, types and functions are defined using Python's syntax, distributed using Python's package system and can call into Python libraries, so @\texttt{lang} is uniquely backwards compatible. We show that active type constructors as implemented in @\texttt{lang} permit practical library implementations of a variety of type system fragments that would normally need to be built into a language. %We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into @\texttt{lang} and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %@\texttt{lang} can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces @\texttt{lang}, a language that aims to satisfy this more comprehensive set of design criteria. To do so, @\texttt{lang} introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of @\texttt{lang} is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Unfortunately, researchers and domain experts designing new abstractions sometimes find it difficult to soundly, completely and naturally implement them as libraries, particularly those that rely on specialized support from a static type system. In these situations, abstraction providers generally develop a new language dialect, increasingly with the aid of tools like language workbenches \cite{erdweg2013state}. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} is problematic at language boundaries: working with components written in several distinct languages often requires programming against their low-level implementation details, rather than their semantics, via a foreign-function interface (FFI) to a common implementation language (e.g. C or the JVM). This is brittle, unnatural and unsafe and can cause performance issues due to necessary checks and conversions at language boundaries. As more languages are used together, more FFIs need to be constructed and traversed.%To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

These issues prevent even those dialects that introduce only a few new constructs from being adopted. A study comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \lstinlinep{java.util.concurrent}, for example, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees \cite{cave2010comparing}. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because using HJ, as a distinct dialect of Java with its own type system, would be problematic in settings where some developers had not adopted it, but needed to interface with code that had. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This appears to be a widespread problem: programmers and development teams often cannot use  the typed abstractions they prefer because they are only available bundled with specialized languages they cannot  adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. %These languages become, in essence, merely vehicles for communicating ideas, rather than useful tools.

@@

Extensible languages have promised to reduce the problems associated with distinct dialects by giving abstraction providers more direct and granular control over a language's syntax and semantics from within, or alongside, libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and so face the classic ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} or Wyvern \cite{TSLs} (which permit only syntax extensions over a fixed semantics) must overcome the same extrinsic barriers as a language like HJ. Second, giving abstraction providers too much control over the language can permit the definition of conflicting  extensions, a problem that is detected ``too late'': when extensions are combined by abstraction clients, who are not necessarily  capable of understanding and fixing the conflict.% Too many constraints can leave it difficult to express real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

@@

Our focus in this paper is to enable the introduction of new static type system fragments as safely composable libraries for an existing and widely-used language, leaving its concrete and abstract syntax fixed. This choice is freeing: questions of syntactic ambiguity no longer need to be addressed. Moreover, many tools operate by exhaustive case analysis over the concrete or abstract syntax of the terms of the language they work with (e.g. syntax highlighters, editor modes, pretty-printers, style checkers, documentation generators). These tools will continue to work. %Indeed, enabling the decentralized introduction of new term constructors is the standard example used when describing the more general problem of enabling the extension of a sum type with new cases as well as new  functions over its cases  \cite{Reynolds75}, prompting Wadler to dub it the \emph{expression problem} \cite{wadler1998expression}.

\paragraph{Approach} Our key insight is that few tools operate by exhaustive case analysis over type constructors. We introduce a mechanism that allows library providers to introduce new type system fragments by defining only new type constructors, in libraries, while leaving the concrete and abstract syntax, and the base semantics of the language, fixed.

To support new operations, the semantics of terms are no longer determined directly based on the identity of their term constructors (the usual ``syntax-directed'' approach). Instead, we delegate to a statically defined function associated with a relevant type constructor. The term constructor's identity only serves to determine how the delegate is assigned. We call user-defined type constructors equipped with these static functions and distributed in libraries \emph{active type constructors}.

As a simple example, consider the concrete term \lstinlinep{e.r} (where \lstinlinep{e} is an expression and \lstinlinep{r} is a static label), which maps to the abstract term $\FF{attr}[\text{``r''}](e)$ (i.e. the term constructor $\FF{attr}$, \emph{indexed} by a static representation of the label and taking one \emph{argument}, $e$). A fragment defining statically typed pairs might see this term as referring to projection of the right component of the pair $e$, while a fragment defining functional record types might want it project out a field labeled \verb|r| from $e$, and an object system might introduce a more complex semantics involving dynamic dispatch. The type of dynamically classified Python values would follow Python's own complex field lookup protocol. Were we to naively ``open up'' the syntax and semantics, these fragments would conflict.

Instead, our semantics will delegate responsibility for assigning a type and a translation to this term to a static function associated with the type constructor of the type recursively synthesized by $e$, the \emph{target} of the operation. Thus, tuples, records, objects, dynamic values and variants thereof can  share this syntax.

As we will detail, terms for which there is no clear ``target'' from which to determine a delegate (e.g. number or list literals) instead either use the type they are being analyzed against (following the approach for composable syntax extensions as supported in Wyvern \cite{TSLs}) or, if not available, a ``default'' delegate determined by the type constructor governing the function in which they appear. Symmetric binary terms (e.g. addition) use a third protocol, which we will discuss. There is always exactly one active type constructor delegated  responsibility over a term and this delegate is stable with respect to additional extension imports, so semantic ambiguities cannot arise by construction (unlike systems that treat the type system as a ``bag of rules''). %This is a generalization of a previous mechanism that we have developed, which deals only with extending the elaboration, but not the typechecking, of the bodies of delimited literals \cite{omar-ecoop14}.

@@

%
Listing \ref{example} shows an example of a well-typed @\texttt{lang} program for which stronger correctness guarantees than Python can statically provide have been established using several type system fragments defined in libraries. As suggested by line \ref{hello-ct}, the top level of every @\texttt{lang} file can be seen as a \emph{compilation script} written directly in Python. @\texttt{lang} requires no modifications to the language (version 2.6+ or 3.0+) or features specific to its main implementation, CPython (so @\texttt{lang} supports alternative implementations like Jython and PyPy). This choice pays off immediately on the first line: @\texttt{lang}'s import mechanism is Python's import mechanism, so Python's build tools (e.g. \lstinlinep{pip}) and package repostories (e.g. \lstinlinep{PyPI}) are directly available for distributing @\texttt{lang} libraries defining type system fragments. Here, \lstinlinep{atlib} is the ``standard library'' but does not benefit from special support in \lstinlinep{atlang} itself.% Nothing in \lstinlinep{atlib} benefits from special support from \lstinlinep{atlang}, the core language package. We will see \lstinlinep{atlang} being used directly below. %@\texttt{lang} can be installed with a single command: \emph{omitted for review}. %@\texttt{lang} is agnostic about the particular design choices that we made in our examples.

@@

\paragraph{Types}
Types are constructed programmatically in the compilation script, always taking the form of a type constructor applied to a type index:  \lstinlinep{tycon[idx]}. Unlike many statically-typed languages, types are constructed programmatically, rather than  declared ``literally'' (e.g. classes in Java, datatypes in ML). %Put another way, 
%@\texttt{lang} supports \emph{static computation at the level of types} and Python is its type-level language. The static language subsumes th%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \lstinlinep{typedef} in C or \lstinlinep{type} in Haskell). 
In this example, we see several useful types being constructed:
\begin{enumerate}
\item On line \ref{decl-Details}, we construct a functional record type with two (immutable) fields, naming it (by assigning it to a static identifier) \lstinlinep{Details}. The field \lstinlinep{amount} has type \lstinlinep{decimal[2]}, which classifies decimal numbers having two decimal places, and \lstinlinep{memo} has type \lstinlinep{dyn}, which comes already constructed, classifying dynamic Python values.  %Here, \lstinlinep{record} and \lstinlinep{decimal} are {indexed type constructors}. 
For convenience, the index for a \lstinlinep{record} type can be provided using borrowed Python  syntax for array slices (e.g. \lstinlinep{a[start:end]}) to approximate conventional notation for type annotations. The field names are simply static strings (and could be computed rather than written literally, e.g. \lstinline{'a' + 'mount'}, to support features like \emph{type providers} \cite{Syme:2013:TIF:2429376.2429378}). 
\item On line \ref{decl-Account-start}, we construct another record type. The field \lstinlinep{name} has type \lstinlinep{string} and the field \lstinlinep{account_num} has a type classifying strings guaranteed statically to be in the regular language specified  by the  regular expression pattern provided as the index (written here as a \emph{raw string literal} \cite{python}). This  fragment, based on recent work, statically tracks the regular language an immutable string is a member of through operations like concatenation and substitution, which is useful for ensuring well-formedness and preventing injection attacks \cite{sanitation-psp14}.%To our knowledge, this type system is not  expressible in other simply typed languages (we do not consider dependently-typed logics without a distinct phase separation).
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line \ref{decl-Transfer}, we construct a simple \emph{prototypic object type} \cite{Lie86}, \lstinlinep{Transfer}, classifying values consisting of a \emph{fore} of type \lstinlinep{Details} and a \emph{prototype} of type \lstinlinep{Account}. %\footnote{We encourage the convention of naming types that are semantic in camelcase and types and type constructors that are structural in nature, like \texttt{dyn} or \texttt{record}, in lowercase.} 
 If a field cannot be found in the fore, the type system will delegate (statically) to the type of the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers to the same account.% differing only in the transaction  details.
\end{enumerate}

\paragraph{Active Type Constructors} Type constructors are Python classes inheriting from \lstinlinep{atlang.Type} (classes here serving as Python's form of open sums). We show portions of \lstinlinep{atlib.fn} in Listings \ref{fn-top} and \ref{fn-stmts}, \lstinlinep{atlib.record} in Listing \ref{record} and \lstinlinep{atlib.string_in} in Listing \ref{string_in}, which we will discuss as we go on. The types in our example are instances of these classes.

\paragraph{Incomplete Types} Incomplete types arise from the application of a type constructor to an index containing an ellipsis (a valid array slice form in Python). The static terms \lstinlinep{fn[Transfer, ...]} and \lstinlinep{fn[...]} seen on lines \ref{log-transfer-start} and \ref{main-start} are incomplete types. We will discuss how the elided portions of an index (e.g. the return types, which need not be provided) will be inferred below.

Incomplete types are not instances of \lstinlinep{atlang.Type} but rather instances of \lstinlinep{atlang.IncompleteType}. An incomplete type is defined by a type constructor (here \lstinlinep{fn}) and an incomplete index (here, \lstinlinep{(Transfer, Ellipsis)}).  The language controls whether the form \lstinlinep{tycon[idx]} results in a complte or incomplete type by overloading the subscript operator on the \lstinlinep{atlang.Type} class object (using Python's \emph{metaclass} mechanism \cite{python}).

@@

 Ascribed function definitions do not share Python's semantics and indeed the underlying Python function is discarded immediately after its abstract syntax and static environment (its closure and the global dictionary of the Python module it was defined in) have been extracted by the compiler using Python's built-in reflection capabilities and \lstinlinep{inspect} and \lstinlinep{ast} packages \cite{python}. The \lstinlinep{ast} package defines Python's term constructors and we will refer to these throughout the paper.
